---
title: "Beluga nucleotide diversity"
output:
  html_document:
    self_contained: yes
date: "2024-05-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8,
  fig.height = 6
)
```

## Setup

Running this will make sure all dependencies of the project will be setup before running anything else:

```{r, eval=FALSE}
install.packages("renv")

renv::restore()
```

## Toy simulations of $Ne_e$ vs $\pi$

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
```

```{r, echo=FALSE}
suppressPackageStartupMessages(library(slendr))

pop <- population("pop", time = 10000, N = 10000)
model1 <- compile_model(pop, generation_time = 1, serialize = FALSE, direction = "backward")
samples <- schedule_sampling(model1, times = c(5000, 4000, 3000, 2000, 1000, 0), list(pop, 50))
p_model1 <- plot_model(model1, samples = samples) + ggtitle("Constant Ne")

pop <- population("pop", time = 10000, N = 10000) %>%
  resize(time = 5000, N = 1000, how = "step")
model2 <- compile_model(pop, generation_time = 1, serialize = FALSE)
samples <- schedule_sampling(model2, times = c(5000, 4000, 3000, 2000, 1000, 0), list(pop, 50))
p_model2 <- plot_model(model2, samples = samples)  + ggtitle("Constant Ne, then a bottleneck")

cowplot::plot_grid(p_model1, p_model2, rel_widths = c(1, 0.85))
```

### Running the simulations

When run, this script will generate the file `pi_simple.rds`:

```
Rscript pi_simple.R
```

### Results

```{r}
pi_simple <- readRDS("pi_simple.rds")
```

```{r}
pi_simple %>% filter(model == "constant Ne") %>%
  ggplot(aes(factor(Ne), diversity, color = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  # geom_jitter(position = position_dodge(width = 0.8), alpha = 0.2, size = 0.75) +
  geom_smooth(aes(group = model), method = "lm", linewidth = 0.5, color = "black", linetype = 2) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "bottom") +
  labs(x = "Ne", y = "nucleotide diversity",
       title = "Expected nucleotide diversity as a function of Ne")
```

```{r}
pi_simple %>% filter(model == "constant Ne" | grepl("1000 gens", model)) %>%
  ggplot(aes(factor(Ne), diversity, color = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  # geom_jitter(position = position_dodge(width = 0.8), alpha = 0.2, size = 0.75) +
  geom_smooth(aes(group = model), method = "lm", linewidth = 0.5, color = "black", linetype = 2) +
  scale_color_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "bottom") +
  labs(x = "Ne", y = "nucleotide diversity",
       title = "Expected nucleotide diversity as a function of Ne")
```

```{r}
pi_simple %>% filter(model == "constant Ne" | grepl("1000 gens", model) | grepl("2000 gens", model)) %>%
  ggplot(aes(factor(Ne), diversity, color = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  # geom_jitter(position = position_dodge(width = 0.8), alpha = 0.2, size = 0.75) +
  geom_smooth(aes(group = model), method = "lm", linewidth = 0.5, color = "black", linetype = 2) +
  scale_color_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "bottom") +
  labs(x = "Ne", y = "nucleotide diversity",
       title = "Expected nucleotide diversity as a function of Ne")
```

```{r}
ggplot(pi_simple, aes(factor(Ne), diversity, color = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  # geom_jitter(position = position_dodge(width = 0.8), alpha = 0.2, size = 0.75) +
  geom_smooth(aes(group = model), method = "lm", linewidth = 0.5, color = "black", linetype = 2) +
  scale_color_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "bottom") +
  labs(x = "Ne", y = "nucleotide diversity",
       title = "Expected nucleotide diversity as a function of Ne")
```

## Simulations of Beluga diversity given a level of hunting

```{r, echo=FALSE, message=FALSE}
library(slendr)
pi_beluga <- readRDS("pi_beluga_50Mb.rds")
plot_model(pi_beluga$model[[20]])
```

```{r, message=FALSE}
library(ggplot2)
library(ggpubr)
library(dplyr)
library(tidyr)
library(cowplot)
library(slendr)
```

### Running the simulations

When run, this script will generate the file `pi_simple.rds`:

```
Rscript pi_beluga.R
```

### Results

```{r}
pi_beluga <- readRDS("pi_beluga_50Mb.rds")# %>% filter(engine == "msprime")

source("plotting.R")
```

```{r}
unique(pi_beluga$Ne_start)
unique(pi_beluga$Ne_hunted)
unique(pi_beluga$census_ratio)
```

```
> unique(pi_beluga$Ne_start)
10000 20000 30000 40000
> unique(pi_beluga$Ne_hunted)
100  250  500  750 1000 1500 2000 3000
> unique(pi_beluga$census_ratio)
0.1 0.2 0.3 0.4 0.5 0.6 0.7 0.8 0.9 1.0
```

```{r}
plot_grid(
  plot_slopes(pi_beluga, "msprime") + labs(subtitle = "msprime engine") + coord_cartesian(ylim = c(-1e-8, 1e-8)),
  plot_slopes(pi_beluga, "SLiM") + labs(subsubtitle ="SLiM engine") + coord_cartesian(ylim = c(-1e-8, 1e-8)),
  nrow = 2
)
```

#### Starting $N_e$ = 40000, $N_e$ hunted = 100, census ratio = 1.0

```{r}
plot_panels(pi_beluga, Ne_start = 40000, Ne_hunted = 100, census_ratio = 1.0, engine = "msprime")
```

<!-- ```{r} -->
<!-- plot_panels(pi_beluga, Ne_start = 40000, Ne_hunted = 100, census_ratio = 1.0, engine = "SLiM") -->
<!-- ``` -->

#### Starting $N_e$ = 40000, $N_e$ hunted = 1000, census ratio = 1.0

```{r}
plot_panels(pi_beluga, Ne_start = 40000, Ne_hunted = 1000, census_ratio = 1.0, engine = "msprime")
```

<!-- ```{r} -->
<!-- plot_panels(pi_beluga, Ne_start = 40000, Ne_hunted = 1000, census_ratio = 1.0, engine = "SLiM") -->
<!-- ``` -->

#### Starting $N_e$ = 10000, $N_e$ hunted = 100, census ratio = 1.0

```{r}
plot_panels(pi_beluga, Ne_start = 10000, Ne_hunted = 100, census_ratio = 1.0, engine = "msprime")
```

<!-- ```{r} -->
<!-- plot_panels(pi_beluga, Ne_start = 10000, Ne_hunted = 100, census_ratio = 1.0, engine = "SLiM") -->
<!-- ``` -->

#### Starting $N_e$ = 10000, $N_e$ hunted = 300, census ratio = 1.0

```{r}
plot_panels(pi_beluga, Ne_start = 10000, Ne_hunted = 250, census_ratio = 1.0, engine = "msprime")
```

<!-- ```{r} -->
<!-- plot_panels(pi_beluga, Ne_start = 10000, Ne_hunted = 250, census_ratio = 1.0, engine = "SLiM") -->
<!-- ``` -->
