---
title: "Beluga nucleotide diversity"
#output:
#  html_document:
#    self_contained: yes
output:
  github_document:
    html_preview: false
date: "2024-05-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  fig.width = 10,
  fig.height = 8
)

set.seed(42)

dir.create("results", showWarnings = FALSE)
```

## Setup

Running this will make sure all dependencies of the project will be setup before running anything else:

```{r, eval=FALSE}
install.packages("renv")

renv::restore()
```

## Toy simulations of $Ne_e$ vs $\pi$

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
```

```{r, echo=FALSE}
suppressPackageStartupMessages(library(slendr))

# pop <- population("pop", time = 10000, N = 10000)
# model1 <- compile_model(pop, generation_time = 1, serialize = FALSE, direction = "backward")
# samples <- schedule_sampling(model1, times = c(5000, 4000, 3000, 2000, 1000, 0), list(pop, 50))
# p_model1 <- plot_model(model1, samples = samples) + ggtitle("Constant Ne")

pop <- population("pop", time = 10000, N = 10000) %>%
  resize(time = 5000, N = 1000, how = "step")
model2 <- compile_model(pop, generation_time = 1, serialize = FALSE)
samples <- schedule_sampling(model2, times = c(5000, 4000, 3000, 2000, 1000, 0), list(pop, 50))
p_model2 <- plot_model(model2, samples = samples)  + ggtitle("Constant Ne, then a bottleneck")
p_model2

# cowplot::plot_grid(p_model1, p_model2, rel_widths = c(1, 0.85))
```

### Running the simulations

When run, this script will generate the file `pi_simple.rds`:

```
Rscript pi_simple.R
```

### Results

```{r}
pi_simple <- readRDS("pi_simple.rds")
```

```{r}
pi_simple %>% filter(model == "constant Ne") %>%
  ggplot(aes(factor(Ne), diversity, color = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  # geom_jitter(position = position_dodge(width = 0.8), alpha = 0.2, size = 0.75) +
  geom_smooth(aes(group = model), method = "lm", linewidth = 0.5, color = "black", linetype = 2) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "bottom") +
  labs(x = "Ne", y = "nucleotide diversity",
       title = "Expected nucleotide diversity as a function of Ne")
```

```{r}
pi_simple %>% filter(model == "constant Ne" | grepl("100 gens", model)) %>%
  ggplot(aes(factor(Ne), diversity, color = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  # geom_jitter(position = position_dodge(width = 0.8), alpha = 0.2, size = 0.75) +
  geom_smooth(aes(group = model), method = "lm", linewidth = 0.5, color = "black", linetype = 2) +
  scale_color_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "bottom") +
  labs(x = "Ne", y = "nucleotide diversity",
       title = "Expected nucleotide diversity as a function of Ne")
```

```{r}
pi_simple %>% filter(model == "constant Ne" | grepl("1000 gens", model)) %>%
  ggplot(aes(factor(Ne), diversity, color = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  # geom_jitter(position = position_dodge(width = 0.8), alpha = 0.2, size = 0.75) +
  geom_smooth(aes(group = model), method = "lm", linewidth = 0.5, color = "black", linetype = 2) +
  scale_color_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "bottom") +
  labs(x = "Ne", y = "nucleotide diversity",
       title = "Expected nucleotide diversity as a function of Ne")
```

```{r}
pi_simple %>% filter(model == "constant Ne" | grepl("1000 gens", model) | grepl("2000 gens", model)) %>%
  ggplot(aes(factor(Ne), diversity, color = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  # geom_jitter(position = position_dodge(width = 0.8), alpha = 0.2, size = 0.75) +
  geom_smooth(aes(group = model), method = "lm", linewidth = 0.5, color = "black", linetype = 2) +
  scale_color_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "bottom") +
  labs(x = "Ne", y = "nucleotide diversity",
       title = "Expected nucleotide diversity as a function of Ne")
```

```{r}
ggplot(pi_simple, aes(factor(Ne), diversity, color = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  # geom_jitter(position = position_dodge(width = 0.8), alpha = 0.2, size = 0.75) +
  geom_smooth(aes(group = model), method = "lm", linewidth = 0.5, color = "black", linetype = 2) +
  scale_color_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "bottom") +
  labs(x = "Ne", y = "nucleotide diversity",
       title = "Expected nucleotide diversity as a function of Ne")
```

## Simulations of Beluga diversity given a level of hunting

```{r, echo=FALSE, message=FALSE}
library(slendr)
pi_beluga <- readRDS("pi_beluga_50Mb.rds")
plot_model(pi_beluga$model[[20]])
```

```{r, message=FALSE}
library(ggplot2)
library(ggpubr)
library(dplyr)
library(tidyr)
library(cowplot)
library(slendr)
```

### Running the simulations

When run, this script will generate the file `pi_simple.rds`:

```
Rscript pi_beluga.R
```

### Results

```{r}
pi_beluga <- readRDS("pi_beluga_50Mb.rds")# %>% filter(engine == "msprime")

source("plotting.R")
```

```{r}
unique(pi_beluga$N_start)
unique(pi_beluga$N_hunted)
unique(pi_beluga$census_ratio)
```

```
> unique(pi_beluga$N_start)
10000 20000 30000 40000
# use: 40000
> unique(pi_beluga$N_hunted)
100  250  500  750 1000 1500 2000 3000
# use: 250  500  1000 2500
> unique(pi_beluga$census_ratio)
0.1 0.2 0.3 0.4 0.5 0.6 0.7 0.8 0.9 1.0
# use: 0.3 0.5 0.7
```

```{r}
plot_grid(
  plot_slopes(pi_beluga, "msprime") + coord_cartesian(ylim = c(-1e-8, 1e-8)),
  plot_slopes(pi_beluga, "SLiM") + coord_cartesian(ylim = c(-1e-8, 1e-8)),
  nrow = 2
)
```

**As a sanity check, all models have been run with both `msprime()` and `slim()` engines of _slendr_. All results are expected to be the same regardless of whether they are run in a coalescent or forward-time setting.**

#### Starting $N_e$ = 40000, $N_e$ hunted = 100, census ratio = 1.0

```{r}
plot_panels(pi_beluga, N_start = 40000, N_hunted = 100, census_ratio = 1.0, engine = "msprime")
```

```{r}
plot_panels(pi_beluga, N_start = 40000, N_hunted = 100, census_ratio = 1.0, engine = "SLiM")
```

#### Starting $N_e$ = 40000, $N_e$ hunted = 1000, census ratio = 1.0

```{r}
plot_panels(pi_beluga, N_start = 40000, N_hunted = 1000, census_ratio = 1.0, engine = "msprime")
```

```{r}
plot_panels(pi_beluga, N_start = 40000, N_hunted = 1000, census_ratio = 1.0, engine = "SLiM")
```

#### Starting $N_e$ = 10000, $N_e$ hunted = 100, census ratio = 1.0

```{r}
plot_panels(pi_beluga, N_start = 10000, N_hunted = 100, census_ratio = 1.0, engine = "msprime")
```

```{r}
plot_panels(pi_beluga, N_start = 10000, N_hunted = 100, census_ratio = 1.0, engine = "SLiM")
```

#### Starting $N_e$ = 10000, $N_e$ hunted = 300, census ratio = 1.0

```{r}
plot_panels(pi_beluga, N_start = 10000, N_hunted = 250, census_ratio = 1.0, engine = "msprime")
```

```{r}
plot_panels(pi_beluga, N_start = 10000, N_hunted = 250, census_ratio = 1.0, engine = "SLiM")
```

## Linear fit metrics across all models

```{r}
markers <- tibble(metric = c("slope", "p.value"), value = c(0, 0.05))

pi_beluga %>%
  unnest(lm) %>% 
  pivot_longer(cols = c("r.squared", "slope", "p.value"), names_to = "metric") %>%
  ggplot(aes(value, color = engine)) +
  geom_density() +
  geom_vline(data = markers, aes(xintercept = value), linetype = "dashed") +
  facet_wrap(metric ~ N_start, scales = "free") +
  ggtitle("p-value, R-squared, and slope vs starting N and hunted N")
```

In all cases, $R^2$ values as well as slope estimates of the linear models are miniscule, indicating basically no effect of time on nucleotide diversity. If there's any effect, it pops up when the starting population size is towards the lower end of the range (10.000) and even then there's practically no measurable effect size.

In fact, we can zoom in on the $p < 0.05$ models and inspect that the slopes and $R^2$ are really mostly just noise:

```{r}
p_p_vs_rsquared <-
  pi_beluga %>%
  unnest(lm) %>%
  ggplot(aes(p.value, r.squared * 100)) +
  geom_point(alpha = 0.5) +
  labs(x = "p-value of the linear regression: nucleotide diversity ~ time",
       y = expression(R^2~~"[% of variance explained]")) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  coord_cartesian(ylim = c(0, 5)) +
  ggtitle("p-value of linear regression vs goodness of fit") +
  facet_wrap(~ engine)

p_p_vs_slope <-
  pi_beluga %>%
  unnest(lm) %>%
  ggplot(aes(p.value, slope, aes(color = N_start))) +
  geom_point(alpha = 0.5) +
  labs(x = "p-value of the linear regression: nucleotide diversity ~ time",
       y = "slope of the linear regression") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_cartesian(ylim = c(-2e-8, 2e-8)) +
  ggtitle("p-value of linear regression vs effect size (slope)") +
  facet_wrap(~ engine)

cowplot::plot_grid(p_p_vs_rsquared, p_p_vs_slope, nrow = 2)
```

_msprime_ vs SLiM sanity check -- if slendr correctly interprets a compiled model, both simulation engines should give the same result in terms of simulated nucleotide diversities:

```{r}
pi_beluga %>%
  unnest(pi) %>%
  select(-model, -lm) %>%
  group_by(N_start, N_hunted, census_ratio, engine) %>%
  summarise(pi = mean(pi)) %>%
  ungroup %>%
  pivot_wider(names_from = "engine", values_from = "pi") %>%
  mutate(N_start = paste("N start =", N_start)) %>%
  # sample_frac(0.1) %>%
  ggplot(aes(SLiM, msprime, color = factor(N_hunted))) +
  geom_point(alpha = 0.8) +
  geom_abline(slope = 1, linetype = "dashed", color = "black") +
  facet_wrap(~ N_start, scales = "free", nrow = 1) +
  theme(legend.position = "bottom") +
  labs(x = expression(paste(pi, " as simulated by SLiM")),
       y = expression(paste(pi, " as simulated by msprime")))
```


### Empirical comparison

```{r}
library(dplyr)
library(readr)

pi_empirical <- read_tsv("pi_empirical.txt") %>%
  pivot_longer(
    cols = everything(),
    names_to = "snapshot", names_prefix = "pi_",
    values_to = "pi"
  ) %>% 
  mutate(snapshot = ifelse(snapshot == "Present", "Contemporary", snapshot),
         snapshot = factor(snapshot,
                           levels = c("Time1", "Time2", "Time3", "Contemporary", "Anadyr", "Bristol", "Cook")))
```

#### Raw results from Figure 3a

```{r}
pi_empirical%>%
ggplot(aes(snapshot, pi, fill = snapshot)) +
  geom_violin()
```

Subset to empirical diversities only in Mackenzie belugas:

```{r}
pi_empirical <- filter(pi_empirical, snapshot %in% c("Time1", "Time2", "Time3", "Contemporary"))
```

Normalize diversities relative to the first time snapshot:

```{r}
pi_empirical$pi_relative <- pi_empirical$pi / mean(pi_empirical[pi_empirical$snapshot == "Time1", ]$pi)
```

```{r}
pi_empirical%>%
ggplot(aes(snapshot, pi_relative, fill = snapshot)) +
  geom_violin()
```

#### Empirical vs simulated diversities

Subset simulation results to those roughly matching the empirical data:

```{r}
pi_simulated <- pi_beluga %>%
  filter(engine == "msprime") %>%
  unnest(pi) %>%
  select(-model, -engine) %>%
  mutate(snapshot = case_when(
    time >= 1290 & time <= 1440 ~ "Snapshot #1",
    time >= 1450 & time <= 1650 ~ "Snapshot #2",
    time >= 1800 & time <= 1870 ~ "Snapshot #3",
    time >= 1950 & time <= 2025 ~ "Contemporary",
    TRUE ~ "other"
  )) %>%
  filter(snapshot != "other") %>%
  mutate(snapshot = factor(snapshot,
                         levels = c("Snapshot #1", "Snapshot #2", "Snapshot #3", "Contemporary"))) %>%
  select(-name, -pop, -time) %>%
  filter( # filter based on Mikkel's ideas for the parameter grid
    N_start == 40000,
    N_hunted %in% c(250, 500, 1000, 2500),
    census_ratio %in% c(0.3, 0.5, 0.7)
  )
```

Normalize nucleotide diversities in each simulation vs the mean in the first respective snapshot:

```{r}
pi_simulated <-
  pi_simulated %>%
  group_by(N_start, N_hunted, census_ratio) %>%
  mutate(mean_pi = mean(pi[snapshot == "Snapshot #1"])) %>%
  ungroup() %>%
  mutate(pi_relative = pi / mean_pi) %>%
  select(-mean_pi)
```

```{r}
p_emp <- ggplot(data = pi_empirical, aes(snapshot, pi_relative)) +
  geom_violin(aes(fill = snapshot)) +
  geom_smooth(method = "lm", aes(group = 1), linetype = "dashed", linewidth = 2) +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.75) +
  coord_cartesian(ylim = c(0.5, 1.5)); p_emp
```

#### Figure #1

```{r}
pi_simulated_labels <-
  pi_simulated %>%
  mutate(
    census_ratio = factor(paste("Ne : census ratio =", census_ratio)),
    N_hunted = paste("N hunted =", N_hunted) %>% factor(., levels = gtools::mixedsort(unique(.))),
  )

p_sim_violins <-
  pi_simulated_labels %>%
  ggplot(aes(snapshot, pi_relative)) +
  geom_violin(aes(fill = snapshot), color = NA, alpha = 0.8) +
  geom_smooth(method = "lm", aes(group = 1), fullrange = TRUE,
              color = "black", se = FALSE, linetype = "dashed") +
  facet_wrap(N_hunted ~ census_ratio, nrow = 3) +
  theme(legend.position = "right", axis.text.x = element_text(hjust = 1, angle = 45)) +
  guides(fill = guide_legend("Time point")) +
  labs(
    x = "",
    y = expression(paste(pi, " relative to the first time point")),
    title = paste("Simulated nucleotide diversity across time assuming\nstarting size of 40.000 individuals")
  ) +
  coord_cartesian(ylim = c(0.8, 1.15)); p_sim_violins

ggsave("results/sim_violins.pdf", p_sim_violins, width = 7, height = 8)
```

Potentially a Figure #2 but the table below is the rare case where a plain text table works better:

```{r}
markers <- tibble(metric = c("slope", "p-value", "R-squared"), value = c(0, 0.05, 0))

lm_metrics <- pi_simulated_labels %>%
  select(-pi, -pi_relative, -snapshot) %>%
  unnest(lm) %>% 
  pivot_longer(cols = c("r.squared", "slope", "p.value"), names_to = "metric") %>%
  mutate(metric = case_when(
    metric == "r.squared" ~ "R-squared",
    metric == "p.value" ~ "p-value",
    metric == "slope" ~ "slope"
  )) %>%
  distinct()

p_metrics <- lm_metrics %>%
  ggplot(aes(value, color = metric)) +
  geom_density() +
  geom_vline(data = markers, aes(xintercept = value), linetype = "dashed") +
  facet_wrap(metric ~ ., scales = "free", nrow = 1) +
  theme(legend.position = "none") +
  ggtitle("Distribution of metrics of the linear fit 'time ~ nucleotide diversity'")
```

#### Table #1

```{r}
lm_table <- pi_simulated_labels %>%
  select(-pi, -pi_relative, -snapshot) %>%
  unnest(lm) %>% 
  mutate(N_hunted = as.numeric(gsub("N hunted = ", "", N_hunted)),
         census_ratio = as.numeric(gsub("Ne : census ratio = ", "", census_ratio))) %>%
  distinct() %>%
  select(N_start, N_hunted, census_ratio, p.value, r.squared, slope) %>%
  mutate(
    p.value = format(p.value, digits = 2, nsmall = 2),
    r.squared = format(r.squared, digits = 1),
    slope = format(slope, digits = 1)
  ) %>%
  arrange(N_start, N_hunted, census_ratio)

library(gt)

gt_table <- lm_table %>%
  gt(auto_align = FALSE) %>%
  cols_label(
    N_start = md("**N start**"),
    N_hunted = md("**N hunted**"),
    census_ratio = md("**Ne : census ratio**"),
    p.value = md("**p-value**"),
    r.squared = md("**R-squared**"),
    slope = md("**slope**")
  )

gtsave(gt_table, "results/table.png")
```

#### Figure #3

```{r}
p_model <- pi_beluga %>%
  filter(N_start == 40000, N_hunted == 500, census_ratio == 1.0, engine == "msprime") %>%
  { .$model[[1]] } %>%
  plot_model(); p_model

ggsave("results/model.pdf", p_model, width = 6, height = 7)
```